Title here

### 2020-07-20 paper/paper.Rmd
move notation to the first intro of PP; modify notation in the 3 algos

The optimisation problem we're interested in is in the context of projection pursuit. Coined by @friedman1974projection, projection pursuit is a method that detects the interesting structure (i.e. clustering, outliers and skewness) of multivariate data via projecting it in lower dimensions. Let $\mathbf{X}_{n \times p}$ be a data matrix, an n-d projection can be seen as a linear transformation: $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis  Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ be an index function that maps the projection basis $\mathbf{A}$ onto a scalar value $I$, this function is commonly known as the projection pursuit index (PPI) function, or index function and can be used to measure the "interestingness" of the projection. A number of indice functions have been proposed in the literature and this includes Legendre index [@friedman1974projection], hermite index [@hall1989polynomial], natural hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection]. [Any literature on holes index? xxx]

@@

Projection pursuit is usually used in conjunction with a tour method called *guided tour*. Tour explores the multivariate data *interactively* via playing a series of projections, that form a *tour path* and guided tour uses the path that is geodesically the shortest. Details of the mathematical construction of a tour path can be found in @buja2005computational. Modified from @buja2005computational, Figure \ref{tour-path} vividly depicts the tour path in guided tour. Guided tour, along with other types of tour, has been implemented in the *tourr* package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

[doubt if I need the following. xxx - maybe an algorithmatic version of guide tour would be helpful for non-expert to understand the method?]
<!-- A tour path includes two major components: a *generator* that generating the target basis that has higher index value than the current basis and an *interpolator* that performing geodesic interpolation between the current and target basis. An illustration modified from [@buja2005computational] vividly depicts the tour path in guided tour. -->

@@

## Optimisation in projection pursuit

[thinking if the following three should be presented as algorithms or plain description is fine. xxx]

Now we formulate the problem in mathmetical notation. Projection pursuit finds the projection basis $\mathbf{A}^J = [\mathbf{a}_1, \cdots, \mathbf{a}_d]$, where $\mathbf{a}_i$ is a $p \times 1$ vector, satisfying the following optimisation problem:

@@

via an iterative search: $\{\mathbf{A}^1, \mathbf{A}^2, \cdots \cdots \mathbf{A}^J\}$. Guided tour interpolates geodesically between each basis found by projection pursuit $A^j$ to form a tour path: $$\{ \mathbf{A}^0, \mathbf{A}^1, {}_{1}\mathbf{A}^1, {}_{2}\mathbf{A}^1,\cdots, {}_{k_1}\mathbf{A}^1, A^2, {}_{1}\mathbf{A}^2, {}_{2}\mathbf{A}^2, \cdots {}_{k_2}\mathbf{A}^2, \cdots, \mathbf{A}^{J-1}, {}_{1}\mathbf{A}^{J-1}, \cdots {}_{k_m}\mathbf{A}^{J-1} ,\mathbf{A}^J \}$$

Where $\mathbf{A}^0$ is the starting basis and $\mathbf{A}^J$ is the final basis that has the largest index value in the searching space. Here we use left subscript to denote the index of the interpolation bases in each iteration i.e. ${}_{2}\mathbf{A}^1$ means the second interpolation basis in the first iteration. The right subcript is reserved for the index of searching points in the optimisation algorithm.

There are three existing methods for optimisating PPI function and we review them below. @posse1995projection proposed a stochastic direct search method, a random search algorithm. Given a current basis $\mathbf{A}^{j}$, a candidate basis $\mathbf{A}^{j +1}_l$ is sampled in the neighbourhood defined by the radius of the p-dimensional sphere, $\alpha$ of the current basis $\mathbf{A}^{j}$ by $$\mathbf{A}^{j+1}_{l} = \mathbf{A}^{j} + \alpha \mathbf{A}_{rand}$$. If the candidate basis has a higher index value than the current basis, it is outputed as the target basis $\mathbf{A}^{j + 1}$ along with other metadata and the current iteration stops. If no basis is found to have higher index value after the maximum number of tries $l_{\max}$, the algorithm stops with nothing outputted. $c$ is the halfing parameter and when $c > 30$, the nieghbourhood parameter $\alpha$ in the next iteration will be reduced by half. The algorithm for a random search is summarised in Algorithm \ref{random-search}.

\newpage
\begin{algorithm}[H]
\SetAlgoLined
  initialisation: $c = 0$, $l = 1$ $l_{\max}$, $\mathbf{A}^{j}$\;
  \While{$l < l_{\max}$}{
    $\mathbf{A}^{j+1}_{l} = \mathbf{A}^{j} + \alpha \mathbf{A}_{rand}$\;
    $I^{j+1}_l = f(\mathbf{A}^{j+1}_l)$\;
    \eIf{$I^{j+1}_l > I^{j}$}{
      $\mathbf{A}^{j+1} = \mathbf{A}^{j+1}_{l}$\;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search: my first algorithm}
  \label{random-search}
\end{algorithm}

@cook1995grand explained the use of a gradient ascent optimisation with the assumption that the index function is continuous and differentiable. Since some indices could be non-differentiable, the computation of derivative is replaced by a pseudo-derivative of evaluating five randomly generated directions in a tiny nearby neighbourhood. Taking a step on the straight derivative direction has been modified to maximise the projection pursuit index along the geodesic direction.

\newpage
\begin{algorithm}[H]
\SetAlgoLined
  initialisation: $l_{\max}$, $\alpha$, $l_{\max}$,$\mathbf{A}^j$\;
  \While{$l < l_{\max}$}{
    Generate $\{A^{(j_1)}: A^{(1)}, \cdots A^{(5)}\}$\;
    Find $A^{J_1}_l$ where $J_1 \in j_1$ such that $\{A^{(J_1)}: \arg \max I(A^{(j_1)})\}$\;
    Denote $F_a = A^j$ and $F_z = A^{(J_1)}$\;
    Conduct singular value decomposition (SVD) for $F_a$ and $F_z$ such that $F_{a}^{T}F_z = V_a\Lambda V_z^T$\;
    Compute $G_a = F_aV_a$, $G_z = F_zV_z$ and $\tau = \arccos(\Lambda)$\;
    ...\;
    Find $A^{(J_2)}$ satisfying $A^{J_2}$ on the interpolation path such that $\{A^{(J_2)}: \arg \max I(A^{(j_2)})\}$\;
    Compute $I^{(J_2)} = f(A^{(J_2)})$, $p_{diff} = (I^{(J_2)} - I^{j})/I^j$\;
      \If{$p_{diff} > 0.001$}{
        $A^{j+1} = A^{(J_2)}$\;
        $I^{j+1} = I^{(J_2)}$\;
        \KwOut{$A^{j+1}$, $I^{j+1}$, $i$ and relevant metadata}
      }
    $i = i + 1$\;
  }
  \caption{searchgeodesic: this part has problem!}
  \label{search_geodesic}
\end{algorithm}

Simulated annealing [@bertsimas1993simulated, @kirkpatrick1983optimization] is a non-derivative procedure based on a non-increasing cooling scheme $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{log(l + 1)}$. The simulated annealing algorithm works as follows. Given a neighbourhood parameter $\alpha$ and a randomly generated orthonormal basis $A_{rand}$, a candidate basis is constructed as $A^{j+1}_l = (1 - \alpha)A^{j} + \alpha A_{rand}$. If the index value of the candidate basis is larger than the one of the current basis, the candidate basis becomes the target basis. If it is smaller, the candidate is accepted with probability $\text{prob} = \min\left\{\exp\left[\frac{I^j - I^{j+1}_l}{T(l)}\right],1\right\}$ where $I^j$ and $I^{j+1}_l$ are the index value of the current and candidate basis respectively. The algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\newpage
\begin{algorithm}[H]
\SetAlgoLined
  initialisation: $T_0$, $\alpha$, $l_{\max}$,$\mathbf{A}^j$\;
  \While{$l < l_{\max}$}{
    $\mathbf{A}^{j+1}_{l} = (1 - \alpha)\mathbf{A}^{j} + \alpha \mathbf{A}_{rand}$ s.t. $\lVert\mathbf{a_i} \rVert = 1$\;
    $I^{j+1}_l = f(\mathbf{A}^{j+1}_{l})$\;
    $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I^{j+1}_l > I^{j}$}{
        $\mathbf{A}^{j+1} = \mathbf{A}^{j+1}_{l}$\;
        $I^{j+1} = I^{j+1}_l$\;
        \KwOut{$\mathbf{A}^{j+1}$, $I^{j+1}$,$l$ and relevant metadata}
      }{
        $\text{prob} = \min\left\{\exp\left[\frac{I^j - I^{j+1}_l}{T(l)}\right],1\right\}$\;
        Draw $\text{rand} \sim \text{Unif(0, 1)}$\;
        \If{$\text{prob} > \text{rand}$}{
          \KwOut{$\mathbf{A}^{j+1}$, $I^{j+1}$, $l$ and relevant metadata}
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Ggplot requires a data frame that contains all the variables to plot and a *global object* is constructed as the data frame supplied to the visual diagnostic plots to better suit the characters of iterative optimisation algorithms. Given an optimisation algorithm, two primary variables of interest are the *decision variable* and the *value of the objective function*. To align with the notation used previously for projection pursuit, we denote the decision variable to be $\mathbf{A}$, which are matrices in projection pursuit algorithms and value of the objective function to be $I$, the index function.

*Iterators* indexes the data collected and has a time series feature that prescribes the its natural ordering. Given the complexity of the algorithm, there could be multiple iterators. The simpliest iterator: *index per observation*, is a unique identifier for each observation in the data. For each level of iteration, we design two indices: *index per iteration* is fixed for each observations in one iteration and has an increment of one once a new iteration starts. *index within iteration* has an increment of one for each observation in an iteartion and starts over from one once a new iteration starts. A sketch of the difference between these three iterators is provided in Figure \ref{iterators}. Depend on the number of iteration level ($L$), the total number of iterators will differ and can be calculated as $1 + 2L$.

@@

There could exist other parameters that are also of our interest but can't be classified as one of the three categories. They are defined under the fourth category: *other parameter of interest: $O$*. In projection pursuit, three other parameters of our interest includes `method`, `alpha` and `info`.  `method` identifies the name of the searching method used and we are interested in comparing the performance between different algorithms under direct search. The neighbourhood parameter `alpha` controls the size of the sampling space and we are interested to understand how the searching space shrinks as the algorithm progresses. `info` labels different stages in the searching process. A sketch of the global object for projection pursuit guided tour is presented in Figure \ref{fig:glb-obj}.

@@

Now we formulate the idea of global object using mathmetical notation. Let $$\mathbf{A}_{j,l,k}$$ denotes the projection matrix in iteration $j$. If it is searching basis produced in the optimisation, then it has a subscript $l$. If it is an interpolation basis, it has a $k$ subscript. For example $\mathbf{A}_{1,2,/}$ means the second searching basis in the first iteration, $\mathbf{A}_{1, /, 2}$ means the second interpolation basis that on the tour path from the current basis $\mathbf{A}^0$ to the target basis found in iteration 1: $\mathbf{A}^1$. The target basis found in each iteration, $\mathbf{A}_{j, l_j, /}$ can also be simplified using a superscript as $\mathbf{A}^j$.

Let $\mathbf{A}^0$ denotes the starting basis. Superscript $j$ is used to denote the number of iteration (index per iteration). Candidate basis are denoted using subscript on the right and interpolation basis are denoted using subscript on the left.  For example, $\mathbf{A}^1_{2}$ means the second candidate basis in the first iteration and ${}_{2}\mathbf{A}^1$ means the second interpolation basis that forms the tour path from the current basis $\mathbf{A}^0$ to the target basis $\mathbf{A}^1$. In iteration two, $\mathbf{A}^1$ becomes the current basis and projection pursuit searches through $\mathbf{A}^2_{1}, \mathbf{A}^2_{2}, \cdots, \mathbf{A}^2_{l_2}$ to find the target basis $\mathbf{A}^2$. The tour path from $\mathbf{A}^1$ to $\mathbf{A}^2$ is constructed using the interpolation bases ${}_{1}\mathbf{A}^2, {}_{2}\mathbf{A}^2, \cdots, {}_{k_2}\mathbf{A}^2$. All the bases together can be arranged in a format that each coloumn represents an iteration as in Equation \ref{eq:matrixA}.

\begin{equation}
\left[
\begin{array}{ccccccc}
\ast & \mathbf{A}_{1,1,\ast} & \mathbf{A}_{2, l_1 +1,\ast} & \cdots & \mathbf{A}_{J,\sum_{i = 1}^{J-1}l_i +1,\ast} & \mathbf{A}_{J+1,\sum_{i = 1}^{J}l_i +1 ,\ast} \\
\ast & \mathbf{A}_{1,2,\ast} & \mathbf{A}_{2,l_1 +2,\ast} & \cdots & \mathbf{A}_{J,\sum_{i = 1}^{J-1}l_i +2,\ast} & \mathbf{A}_{J+1,\sum_{i = 1}^{J}l_i +2, \ast} \\
\vdots & \vdots & \vdots & \ddots & \vdots &\vdots   \\
\ast & \mathbf{A}_{1, l, \ast} & \mathbf{A}_{2, l_1 + l, \ast} & \cdots & \mathbf{A}_{J+1, \sum_{i = 1}^{J-1}l_i + l, \ast}  & \mathbf{A}_{J + 1, \sum_{i = 1}^{J}l_i + l, \ast}\\
\vdots & \vdots & \vdots & \ddots & \vdots &\vdots   \\
\ast & \mathbf{A}_{1, l_1-1, \ast} & \mathbf{A}_{2, l_1 + l_2-1, \ast} & \cdots & \mathbf{A}_{J, \sum_{i = 1}^{J-1}l_i -1, \ast} & \mathbf{A}_{J+1, \sum_{i = 1}^{J+1} l_i-1, \ast}\\
\mathbf{A}_{0,\ast,\ast} & \mathbf{A}_{1, l_1, \ast} & \mathbf{A}_{2, l_1 + l_2, \ast} & \cdots & \mathbf{A}_{J, \sum_{i = 1}^{J}l_i , \ast} & \ast\\
\hline
\ast &\mathbf{A}_{1, \ast, 1} & \mathbf{A}_{2, \ast, k_1 + 1} & \cdots &\mathbf{A}_{J, \ast,\sum_{i = 1}^{J-1}k_i + 1}  & \ast\\
\ast &\mathbf{A}_{1, \ast, 2} & \mathbf{A}_{2, \ast, k_1 + 2} & \cdots &\mathbf{A}_{J, \ast, \sum_{i = 1}^{J-1}k_i +2}  & \ast \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\ast &\mathbf{A}_{1, \ast, k} & \mathbf{A}_{2, \ast, k_1 + k} & \cdots &\mathbf{A}_{J, \ast, \sum_{i = 1}^{J-1}k_i +k}  & \ast \\
\vdots & \vdots & \vdots & \ddots & \vdots   & \vdots \\
\ast &\mathbf{A}_{1, \ast, k_1} & \mathbf{A}_{2, \ast, k_1 + k_2} & \cdots &\mathbf{A}_{J, \ast,\sum_{i = 1}^{J}k_i} & \ast \\
\end{array}
\right]
= 
\left[
\begin{array}{c}
\vphantom{A^J_1}\text{candidate basis} \\
\vphantom{A^J_2}\text{candidate basis} \\
\vphantom{\vdots} \vdots \\
\vphantom{A^J_2}\text{candidate basis} \\
\vphantom{\vdots} \vdots \\
\vphantom{A^J_2}\text{candidate basis} \\
\vphantom{A^J}\text{current/target basis} \\
\hline
\vphantom{\text{NULL}}\text{interpolation basis} \\
\vphantom{\text{NULL}}\text{interpolation basis} \\
\vphantom{\text{NULL}}\vdots \\
\vphantom{\text{NULL}}\text{interpolation basis} \\
\vphantom{\vdots} \vdots \\
\vphantom{\text{NULL}}\text{interpolation basis} \\
\end{array}
\right]
\label{eq:matrixA}
\end{equation}

When aggregated with information of index value and other interesting parameter, it is easier to write the basis as a column vector denoting by a single subscript by their natural ordering i.e. all the bases in iteration 1 along with the starting basis as shown in Equation \ref{eq:vectorA}, can be denoted as $\mathbf{A}_0, \mathbf{A}_1, \cdots, \mathbf{A}_n$ where $n = 1 + l_1 + k_1$. The full data structure can thus be presneted as in Equation \ref{eq:data-structure}.

>>>>

The optimisation problem we're interested in is in the context of projection pursuit. Coined by @friedman1974projection, projection pursuit is a method that detects the interesting structure (i.e. clustering, outliers and skewness) of multivariate data via projecting it in lower dimensions. Let $\mathbf{X}_{n \times p}$ be a data matrix, an n-d projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis. Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ be an index function that maps the projection basis $\mathbf{A}$ onto a scalar value $I$, this function is commonly known as the projection pursuit index (PPI) function, or the index function and can be used to measure the "interestingness" of the projection. A number of indice functions have been proposed in the literature including Legendre index [@friedman1974projection], hermite index [@hall1989polynomial], natural hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection].

@@

Projection pursuit is usually used in conjunction with a tour method called *guided tour*. Tour explores the multivariate data *interactively* via playing a series of projections, that form a *tour path* and guided tour uses the path that is geodesically the shortest. Details of the mathematical construction of a tour path can be found in @buja2005computational.  Guided tour, along with other types of tour, has been implemented in the *tourr* package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

<!-- A tour path includes two major components: a *generator* that generating the target basis that has higher index value than the current basis and an *interpolator* that performing geodesic interpolation between the current and target basis. An illustration modified from [@buja2005computational] vividly depicts the tour path in guided tour. -->

@@

For a projection basis, let the three-subscript notation $$\mathbf{A}_{jlk}$$ represent the projection matrix in iteration $j$ with either a searching index $l$ or an interpolation index $k$. The placeholder $*$ will be the substitue for the non-applicable index. To give two examples, $\mathbf{A}_{12*}$ denotes the second searching basis in the first iteration; $\mathbf{A}_{1*2}$ denotes the second interpolation basis in the first iteration.

If we are being pedantic, all the target bases, except the starting basis, should have a searching index since it is the last candidate basis that successfully allows the optimisation algorithm to end and output a new basis for interpolation. If we denotes the length of each search (number of basis searched in each iteration) as $l_1, l_2, \cdots, l_J$ and the length of each interpolation is $k_1, k_2, \cdots, k_J$ , then the searching index for target basis in each iteration will become $l_1, l_1 + l_2, \cdots, l_1 +l_2+\cdots l_J$. To avoid this complex notation involving opteration in the subscript, we simplify it using a superscript as $$\mathbf{A}^{0}, \mathbf{A}^{1}, \mathbf{A}^{2}, \cdots, \mathbf{A}^{J}$$ where $\mathbf{A}^{0}$ is the starting basis. Notice that in the case when the complex notation can't be avoided, we shall separate each index using a comma and  write the index in brackets if necessary. Using the notation above, all the bases on the interpolation path (including target and interpolation bases) can be written as in Equation \ref{eq:interp-path}.

\begin{equation}
\{ 
\mathbf{A}^0, 
\mathbf{A}^1, 
\mathbf{A}_{1 \ast 1}, 
\mathbf{A}_{1 \ast 2},
\cdots, 
\mathbf{A}^2, 
\mathbf{A}_{2 \ast 1}, 
\mathbf{A}_{2 \ast 2}, 
\cdots, 
\mathbf{A}^J,
\mathbf{A}_{J \ast 1}, 
\mathbf{A}_{J \ast 2}, 
\cdots, 
\mathbf{A}_{J \ast K},
\}
\label{eq:interp-path}
\end{equation}

where $K$ denotes the last interpolating index. A visual representation of the interpolation path ( for iteration one) modified from @buja2005computational is shown in Figure \ref{tour-path}. If we use a dot to represent a plane in the searching space, the searching points along with the current and target basis in iteration one can sketched in Figure \ref{searching-points}. Combine the two above, we can write all the bases in the first iteration (including the starting basis) in its natural ordering as in Equation \ref{eq:vectorA}.

\begin{equation}
\{
\mathbf{A}^{0}, 
\mathbf{A}_{1 1\ast}, 
\mathbf{A}_{1 2\ast}, 
\cdots, 
\mathbf{A}^{1},
\mathbf{A}_{1\ast 1}, 
\mathbf{A}_{1\ast 2},  
\cdots, 
\mathbf{A}_{1\ast k_1}
\}
\label{eq:vectorA}
\end{equation}

```{r searching-points, out.width="100%", out.height="60%", fig.cap="\\label{searching-points}An illustration of the tour path"}
knitr::include_graphics(path = here::here("figures/tour_path_increasing.png"))
```

## Optimisation problem and existing algorithms

Now we begin to formulate the optimisation problem. Given a randomly generated starting basis $\mathbf{A}^0$, projection pursuit finds the final projection basis $\mathbf{A}^J = [\mathbf{a}_1, \cdots, \mathbf{a}_d]$, where $\mathbf{a}_i$ is a $p \times 1$ vector, satisfies the following optimisation problem:

@@

via an iterative search of target basis: $\{\mathbf{A}^0, \mathbf{A}^1, \mathbf{A}^2, \cdots \cdots \mathbf{A}^J\}$.

There are three existing methods for optimisating PPI function and we review them below.

@posse1995projection proposed a stochastic direct search method, a random search algorithm. Given a current basis $\mathbf{A}^{j-1}$, a candidate basis $\mathbf{A}_{jl*}$ is sampled in the neighbourhood defined by the radius of the p-dimensional sphere, $\alpha$ of the current basis $\mathbf{A}^{j}$ by $\mathbf{A}_{jl*} = \mathbf{A}^{j-1} + \alpha \mathbf{A}_{rand}$. If the candidate basis has a higher index value than the current basis, it is outputed as the target basis $\mathbf{A}^{j}$ along with other metadata and the current iteration stops. If no basis is found to have higher index value after the maximum number of tries $l_{\max}$, the algorithm stops with nothing outputted. $c$ is the halfing parameter and when $c > 30$, the nieghbourhood parameter $\alpha$ in the next iteration will be reduced by half. The algorithm for a random search is summarised in Algorithm \ref{random-search}.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{jl*} = \mathbf{A}^{j-1} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{j, l,*}$ is orthonormal\;
    Compute $I_{jl*}  = f(\mathbf{A}_{jl*})$\;
    \eIf{$I_{jl*} > I^{j-1}$}{
      $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

@cook1995grand explained the use of a gradient ascent optimisation with the assumption that the index function is continuous and differentiable. Since some indices could be non-differentiable, the computation of derivative is replaced by a pseudo-derivative of evaluating five randomly generated directions in a tiny nearby neighbourhood. Taking a step on the straight derivative direction has been modified to maximise the projection pursuit index along the geodesic direction. See Algorithm \ref{search-geodesic} for a summarisation of the steps.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate ten bases in five random directions: $\mathbf{A}_{jl*}: \mathbf{A}_{j,(l + 9), *}$ within a small neighbourhood $\text{dist}$\;
    Find the direction with the largest index value: $\mathbf{A}_{j, l_{d}, *}$\;
    Construct the geodesic from $\mathbf{A}^{j-1}$ to $\mathbf{A}_{j, l_{d}, *}$\;
    Find $\mathbf{A}_{j,l_g, *}$ on the geodesic that has the largest index value \;
    Compute $I_{j,l_g, *} = f(\mathbf{A}_{j,l_g, *})$, $p_{\text{diff}} = (I_{j,l_g, *} - I^{j-1})/I_{j,l_g, *}$\;
      \If{$p_{\text{diff}} > 0.001$}{
        $A^{j} = A_{j,l_g, *}$, $I^{j} = I_{j,l_g, *}$\;
      }
    $l = l + 1$\;
  }
  \caption{search geodesic}
  \label{search-geodesic}
\end{algorithm}

Simulated annealing [@bertsimas1993simulated, @kirkpatrick1983optimization] is a non-derivative procedure based on a non-increasing cooling scheme $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{log(l + 1)}$. The simulated annealing algorithm works as follows. Given a neighbourhood parameter $\alpha$ and a randomly generated orthonormal basis $A_{rand}$, a candidate basis is constructed as $\mathbf{A}_{jl*} = (1 - \alpha)\mathbf{A}^{j} + \alpha \mathbf{A}_{rand}$. If the index value of the candidate basis is larger than the one of the current basis, the candidate basis becomes the target basis. If it is smaller, the candidate is accepted with probability $P= \min\left\{\exp\left[\frac{I^{j-1} - I_{jl*}}{T(l)}\right],1\right\}$ where $I^{j-1}$ and $I_{jl*}$ are the index value of the current and candidate basis respectively. The algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{jl*} = (1 - \alpha)\mathbf{A}^{j-1} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{jl*}$ is orthonormal \;
    Commpute $I_{jl*} = f(\mathbf{A}_{jl*})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{jl*} > I^{j-1}$}{
        $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I^{j-1} - I_{jl*}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
          $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Ggplot requires a data frame that contains all the variables to plot and a *global object* is constructed as the data frame supplied to the visual diagnostic plots to better suit the characters of iterative optimisation algorithms. Given an optimisation algorithm, two primary variables of interest are the *decision variable: $\mathbf{A}$* and the *value of the objective function: $I$*.

*Iterators* indexes the data collected and has a time series feature that prescribes the its natural ordering of the decision varible in the searching. The simpliest iterator, *index per observation: $t$*, is a unique identifier for each observation in the data. For each level of iteration, we design two types of indices: *index per iteration* is fixed for each observations in one iteration and has an increment of one once a new iteration starts. *index within iteration* has an increment of one for each observation in an iteartion and starts over from one once a new iteration starts. A sketch of the difference between these three iterators is provided in Figure \ref{iterators}.

@@

There could exist other parameters that are also of our interest but can't be classified as one of the three categories. They are defined under the fourth category: *other parameter of interest: $S$*. In projection pursuit, three other parameters of our interest includes `method`, `alpha` and `info`.  `method` identifies the name of the searching method used and we are interested in comparing the performance between different algorithms under direct search. The neighbourhood parameter `alpha` controls the size of the sampling space and we are interested to understand how the searching space shrinks as the algorithm progresses. `info` labels different stages in the searching process. A sketch of the global object for projection pursuit guided tour is presented in Figure \ref{fig:glb-obj}.

@@

<!-- \begin{equation} -->
<!-- \left[ -->
<!-- \begin{array}{ccccccc} -->
<!-- \ast & \mathbf{A}_{1,1,\ast} & \mathbf{A}_{2, l_1 +1,\ast} & \cdots & \mathbf{A}_{J,\sum_{i = 1}^{J-1}l_i +1,\ast} & \mathbf{A}_{J+1,\sum_{i = 1}^{J}l_i +1 ,\ast} \\ -->
<!-- \ast & \mathbf{A}_{1,2,\ast} & \mathbf{A}_{2,l_1 +2,\ast} & \cdots & \mathbf{A}_{J,\sum_{i = 1}^{J-1}l_i +2,\ast} & \mathbf{A}_{J+1,\sum_{i = 1}^{J}l_i +2, \ast} \\ -->
<!-- \vdots & \vdots & \vdots & \ddots & \vdots &\vdots   \\ -->
<!-- \ast & \mathbf{A}_{1, l, \ast} & \mathbf{A}_{2, l_1 + l, \ast} & \cdots & \mathbf{A}_{J+1, \sum_{i = 1}^{J-1}l_i + l, \ast}  & \mathbf{A}_{J + 1, \sum_{i = 1}^{J}l_i + l, \ast}\\ -->
<!-- \vdots & \vdots & \vdots & \ddots & \vdots &\vdots   \\ -->
<!-- \ast & \mathbf{A}_{1, l_1-1, \ast} & \mathbf{A}_{2, l_1 + l_2-1, \ast} & \cdots & \mathbf{A}_{J, \sum_{i = 1}^{J-1}l_i -1, \ast} & \mathbf{A}_{J+1, \sum_{i = 1}^{J+1} l_i-1, \ast}\\ -->
<!-- \mathbf{A}_{0,\ast,\ast} & \mathbf{A}_{1, l_1, \ast} & \mathbf{A}_{2, l_1 + l_2, \ast} & \cdots & \mathbf{A}_{J, \sum_{i = 1}^{J}l_i , \ast} & \ast\\ -->
<!-- \hline -->
<!-- \ast &\mathbf{A}_{1, \ast, 1} & \mathbf{A}_{2, \ast, k_1 + 1} & \cdots &\mathbf{A}_{J, \ast,\sum_{i = 1}^{J-1}k_i + 1}  & \ast\\ -->
<!-- \ast &\mathbf{A}_{1, \ast, 2} & \mathbf{A}_{2, \ast, k_1 + 2} & \cdots &\mathbf{A}_{J, \ast, \sum_{i = 1}^{J-1}k_i +2}  & \ast \\ -->
<!-- \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\ -->
<!-- \ast &\mathbf{A}_{1, \ast, k} & \mathbf{A}_{2, \ast, k_1 + k} & \cdots &\mathbf{A}_{J, \ast, \sum_{i = 1}^{J-1}k_i +k}  & \ast \\ -->
<!-- \vdots & \vdots & \vdots & \ddots & \vdots   & \vdots \\ -->
<!-- \ast &\mathbf{A}_{1, \ast, k_1} & \mathbf{A}_{2, \ast, k_1 + k_2} & \cdots &\mathbf{A}_{J, \ast,\sum_{i = 1}^{J}k_i} & \ast \\ -->
<!-- \end{array} -->
<!-- \right] -->
<!-- =  -->
<!-- \left[ -->
<!-- \begin{array}{c} -->
<!-- \vphantom{A^J_1}\text{candidate basis} \\ -->
<!-- \vphantom{A^J_2}\text{candidate basis} \\ -->
<!-- \vphantom{\vdots} \vdots \\ -->
<!-- \vphantom{A^J_2}\text{candidate basis} \\ -->
<!-- \vphantom{\vdots} \vdots \\ -->
<!-- \vphantom{A^J_2}\text{candidate basis} \\ -->
<!-- \vphantom{A^J}\text{current/target basis} \\ -->
<!-- \hline -->
<!-- \vphantom{\text{NULL}}\text{interpolation basis} \\ -->
<!-- \vphantom{\text{NULL}}\text{interpolation basis} \\ -->
<!-- \vphantom{\text{NULL}}\vdots \\ -->
<!-- \vphantom{\text{NULL}}\text{interpolation basis} \\ -->
<!-- \vphantom{\vdots} \vdots \\ -->
<!-- \vphantom{\text{NULL}}\text{interpolation basis} \\ -->
<!-- \end{array} -->
<!-- \right] -->
<!-- \label{eq:matrixA} -->
<!-- \end{equation} -->

To write down the notation for the data structure, it is easier to write the bases as a column vector denoting by a single subscript by their natural ordering:

\begin{equation}
\{
\mathbf{A}_0, 
\mathbf{A}_1, 
\mathbf{A}_2, 
\cdots,
\mathbf{A}_T
\label{vector-basis}
\}
\end{equation}

where $T$ is the total number of plane serached by the optimisation algorithm including all the target planes, all the interpolating planes and all the candidate planes but NOT the starting plane. The full data structure can thus be shown as in Equation \ref{eq:data-structure}.
